A Large-Scale Study of the Use of Eval in R

### 2020-08-28 inst/paper/main.tex
Fix number of call sites per traced packages
Typos

Table~\ref{table:site-package-summary} shows the number of call sites in
CRAN packages. 61 packages have a single call site, and one package, \emph{VGAM}, has \MaxCallSitePackage
call sites of \eval. \NoCallSitePackages packages have no calls to \eval. The core packages
have \NbCoreEvalCallSites call sites of \eval and Kaggle has none. These numbers are
under-approximations as \eval may be called reflectively, through aliases,
or passed to a higher-order function.

@@

In our corpus, we observe 5.2\% cases where \eval directly evaluates output of
\c{substitute}, 0.7\% cases where output of \c{parse} and its variants is read
directly and only 704 cases for \c{expression}. Majority of expression objects
consumed by \eval are generated by their transitive uses by other functions.

The pattern \c{eval(parse(...))} can be used for dynamic code loading. This
forms the core of \c{source} and \c{sys.source} functions in R that are commonly
used for loading code in R files in interactive settings. We investigated the
number of cases in which the output of \c{parse} and its variants is passed to
\eval, directly or transitively by tainting their output. This corresponds to \PercentParsedCallSites of the total \eval call sites and
\PercentParsedEvals of the eval calls. We observed that
very few of the eval call(\NbParseFilesRnd in total) consume the result of calling \c{parse} on
a file. Most of the eval calls consume the result of calling
\c{parse} on a string.  We also identified one
function in core R, \c{invokeRestartInteractively} that prompts the user for
input, parses it, and passes it to \eval.

@@

In the corpus we observe \AllWritesRnd writes to variables of which
\EvalWritesRnd writes happen inside \eval. However, all writes are not
dangerous. Only writes to environments not local to the computation spawned
by \eval are side-effecting. These writes outlive the computation and hence are
visible outside the computation. The remaining writes are local to the
computation. We observe that \EvalSideEffectingWritesRnd writes inside evals are
side-effecting. This is only \EvalSideEffectingWritesEvalPerc of all variable
writes inside \eval and \EvalSideEffectingWritesAllPerc of all variable writes
in the corpus.

@@

This argument can also be \c{NULL}, a \c{List} or a data frame. This happens in
3.5\% of the cases.
\c{eval} copies the fields of the list or data frame and creates bindings for
them in a new environment. This pattern is used to evaluate formulas which can
directly refer to the fields of the data.
The \c{envir} argument can also be a number $n$. It means that the environment in
which the expressions is evaluated will be the result of \c{sys.call(n)} where
$n$ refers to the $n$-th stack frame.

@@

\begin{table}[htbp]
    \centering
    \begin{tabular}{c|c|c|c}
        \hline
        \multicolumn{2}{c|}{Core} & \multicolumn{2}{|c}{Packages}  \\
        \hline
        Environment Class & Eval Calls \% & Environment Class & Eval Calls \% \\
        \hline
        \CoreEnvClassA & \CoreProportionA & \PackagesEnvClassA &  \PackagesProportionA\\
        \CoreEnvClassB & \CoreProportionB & \PackagesEnvClassB &  \PackagesProportionB\\
        \CoreEnvClassC & \CoreProportionC & \PackagesEnvClassC &  \PackagesProportionC\\
        \CoreEnvClassD & \CoreProportionD & \PackagesEnvClassD &  \PackagesProportionD\\
        \CoreEnvClassE & \CoreProportionE & \PackagesEnvClassE &  \PackagesProportionE\\
        \hline
    \end{tabular} \label{tab:environments}
\caption{Environments in terms of eval calls.}
\end{table}

We observe that a disproportionately high core R \eval calls access the
caller's caller's environment. This is because many core R functions call
functions that pass the result of \c{parent.frame()} to an \eval.
A disproportionately high calls to \eval happen in an extended
top-level environment. This can be explained by the fact that many packages
evaluate code passed from the user's workspace in the top-level environment to
access the bindings.

>>>>

Table~\ref{table:site-package-summary} shows the number of call sites in
CRAN packages. \OneCallSitesEnrich packages have a single call site, and one package, \emph{metafor}, has \MaxCallSitesEnrich
call sites of \eval. \NoCallSitesEnrich packages have no calls to \eval. The core packages
have \NbCoreEvalCallSites call sites of \eval and Kaggle has none. These numbers are
under-approximations as \eval may be called reflectively, through aliases,
or passed to a higher-order function.

@@

In our corpus, we observe 5.2\% cases where \eval directly evaluates the output of
\c{substitute}, 0.7\% cases where output of \c{parse} and its variants is read
directly and only 704 cases for \c{expression}. Majority of expression objects
consumed by \eval are generated by their transitive uses by other functions.

The pattern \c{eval(parse(...))} can be used for dynamic code loading. This
forms the core of \c{source} and \c{sys.source} functions in R that are commonly
used for loading code in R files in interactive settings. We investigated the
number of cases in which the output of \c{parse} and its variants is passed to
\eval, directly or transitively by tainting their output. This corresponds to \PercentParsedCallSites of the total \eval call sites and
\PercentParsedEvals of the eval calls. We observed that
very few of the eval call (\NbParseFilesRnd in total) consume the result of calling \c{parse} on
a file. Most of the eval calls consume the result of calling
\c{parse} on a string.  We also identified one
function in core R, \c{invokeRestartInteractively} that prompts the user for
input, parses it, and passes it to \eval.

@@

In the corpus we observe \AllWritesRnd writes to variables of which
\EvalWritesRnd writes happen inside \eval. However, all writes are not
dangerous. Only writes to environments not local to the computation spawned
by \eval are side-effecting. These writes outlive the computation and hence are
visible outside it. The remaining writes are local to the
computation. We observe that \EvalSideEffectingWritesRnd writes inside evals are
side-effecting. This is only \EvalSideEffectingWritesEvalPerc of all variable
writes inside \eval and \EvalSideEffectingWritesAllPerc of all variable writes
in the corpus.

@@

This argument can also be \c{NULL}, a \c{List} or a data frame. This happens in
3.5\% of the cases: \c{eval} copies the fields of the list or data frame and creates bindings for
them in a new environment. This pattern is used to evaluate formulas which can
directly refer to the fields of the data.
The \c{envir} argument can also be a number $n$. It means that the environment in
which the expressions is evaluated will be the result of \c{sys.call(n)} where
$n$ refers to the $n$-th stack frame.

@@

\begin{table}[htbp] \resizebox{\columnwidth}{!}{%
    \centering
    \begin{tabular}{c|c|c|c}
        \hline
        \multicolumn{2}{c|}{Core} & \multicolumn{2}{|c}{Packages}  \\
        \hline
        Environment Class & Eval Calls \% & Environment Class & Eval Calls \% \\
        \hline
        \CoreEnvClassA & \CoreProportionA & \PackagesEnvClassA &  \PackagesProportionA\\
        \CoreEnvClassB & \CoreProportionB & \PackagesEnvClassB &  \PackagesProportionB\\
        \CoreEnvClassC & \CoreProportionC & \PackagesEnvClassC &  \PackagesProportionC\\
        \CoreEnvClassD & \CoreProportionD & \PackagesEnvClassD &  \PackagesProportionD\\
        \CoreEnvClassE & \CoreProportionE & \PackagesEnvClassE &  \PackagesProportionE\\
        \hline
    \end{tabular}} \label{tab:environments}
\caption{Environments in terms of eval calls.}
\end{table}

We observe that a disproportionately high number of core R \eval calls access the
caller's caller's environment. This is because many core R functions call
functions that pass the result of \c{parent.frame()} to an \eval.
A disproportionately high number of calls to \eval happens in an extended
top-level environment. This can be explained by the fact that many packages
evaluate code passed from the user's workspace in the top-level environment to
access the bindings.
